market-justice-preferences

### 2024-06-28 05-discussion.qmd
revision JC

Hypothesis $H_4$ held that students from schools with higher school achievement would show stronger preferences for market justice, following the argument that such school could foster meritocratic attitudes by showing a larger promotion of talent and effort in order to get better results. Nevertheless, the results show quite the opposite: schools with better academic results (as measured by a national standardized test) show in average less preferences for market justice. This finding is one of the most strongest and consistent throughout the models and it is relevant given the centrality of school performance in educational institutions. One possible explanation of this link is that students in high-achieving schools may be more exposed to discussions and to a deeper understanding about inequality and social justice, leading to a critical perspective towards the weight of market mechanism in the access to social services.

Hypotheses $H_5$ and $H_6$ proposed that socioeconomic status would moderate the effect of meritocratic perceptions on market justice preferences. Hypothesis $H_5$ refers to family socioeconomic status, whereby the results show a negative interaction with two of the three perceptual variables at the society level. This means that the positive impact of the perception of social meritocracy on market justice preferences have a lower effect in students with parents with higher education and also in those with more books at home. Something similar occurs with hypothesis $H_6$, where now school level status moderates the effect of meritocracy variables, which in multilevel framework si referred as to cross-level interaction. The negative interaction in this context means that in schools of low socioeconomic levels, the positive effect of the social meritocracy variables on market justice preferences is stronger. These relations indicate that, although those who perceive society as meritocratic have greater preferences towards market justice, this influence is lower among those with a higher socioeconomic level. This finding highlights the complex interaction between socioeconomic status, school context, meritocratic beliefs, and preferences toward market justice. The moderating role of meritocratic perceptions suggests that students' views on social justice are shaped by their socioeconomic background and beliefs about how society rewards effort and talent. This underlines the importance of addressing structural inequalities and individual beliefs in educational and social policies.

>>>>

Hypothesis $H_4$ held that students from schools with higher school achievement would show stronger preferences for market justice, following the argument that such schools could foster meritocratic attitudes by showing a larger promotion of talent and effort in order to get better results. Nevertheless, the results show quite the opposite: schools with better academic results (as measured by a national standardized test) show in average less preferences for market justice. This finding is one of the most strongest and consistent throughout the models and it is relevant given the centrality of school performance in educational institutions. One possible explanation of this link is that students in high-achieving schools may be more exposed to discussions and a deeper understanding of inequality and social justice, leading to a critical perspective toward the weight of market mechanisms in the access to social services.

Hypotheses $H_5$ and $H_6$ proposed that socioeconomic status would moderate the effect of meritocratic perceptions on market justice preferences. Hypothesis $H_5$ refers to family socioeconomic status, whereby the results show a negative interaction with two of the three perceptual variables at the societal level. This means that the positive impact of the perception of social meritocracy on market justice preferences had a lower effect on students with parents with higher education and also on those with more books at home. Something similar occurs with hypothesis $H_6$, where now school-level status moderates the effect of meritocracy variables, which in multilevel framework is referred as to cross-level interaction. The negative interaction in this context means that in schools of low socioeconomic levels, the positive effect of the social meritocracy variables on market justice preferences is stronger. These relations indicate that, although those who perceive society as meritocratic have greater preferences towards market justice, this influence is lower among those with a higher socioeconomic level. This finding highlights the complex interaction between socioeconomic status, school context, meritocratic beliefs, and preferences toward market justice. The moderating role of meritocratic perceptions suggests that students' views on social justice are shaped by their socioeconomic background and beliefs about how society rewards effort and talent. This underlines the importance of addressing structural inequalities and individual beliefs in educational and social policies.
